---
title: "Fitting Multiple Motif Types"
author: "Andrew G. Brown"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting Multiple Motif Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(soilmotif)
```

# Introduction

This vignette demonstrates how to fit multiple soil depth motif types to soil property data and select the best-fitting model. We'll cover:

- Preparing soil profile data for motif fitting
- Fitting individual motifs with optimization
- Automated model selection across all motif types
- Interpreting and visualizing results

# Data Preparation

Soil motif fitting requires continuous depth profiles. The `sm_prepare()` function handles conversion from discrete horizon data to continuous profiles using different interpolation methods.

```{r data_prep}
# Example horizon data
horizon_data <- data.frame(
  id = c("P1", "P1", "P1", "P1"),
  top = c(0, 15, 45, 90),
  bottom = c(15, 45, 90, 120),
  clay = c(8, 25, 35, 42)
)

# Convert to continuous profile
profiles <- sm_prepare(horizon_data, property = "clay", method = "linear")
clay_profile <- profiles$clay$P1

# Plot the profile
depths <- 1:200
plot(clay_profile, depths, ylim = c(200, 1), type = "l", 
     xlab = "Clay Content (%)", ylab = "Depth (cm)",
     main = "Prepared Clay Profile")
```

# Fitting Individual Motifs

You can fit individual motif types using `sm_optim()`. Each motif type has different parameters:

```{r individual_fits}
# Fit different motif types
sigmoid_fit <- sm_optim(clay_profile, c(50, 100), FUN = sm_shape_wetting_front)
peak_fit <- sm_optim(clay_profile, c(80, 20), FUN = sm_shape_peak)
gradational_fit <- sm_optim(clay_profile, c(10, 40), FUN = sm_shape_gradational)

# Plot comparisons
plot(clay_profile, depths, ylim = c(200, 1), type = "l", lwd = 2,
     xlab = "Clay Content (%)", ylab = "Depth (cm)",
     main = "Individual Motif Fits")
lines(sigmoid_fit, depths, col = "blue", lty = 2, lwd = 2)
lines(peak_fit, depths, col = "red", lty = 3, lwd = 2)
lines(gradational_fit, depths, col = "green", lty = 4, lwd = 2)
legend("bottomright", 
       c("Observed", "Wetting Front", "Peak", "Gradational"),
       lty = c(1, 2, 3, 4), col = c("black", "blue", "red", "green"), lwd = 2)
```

# Automated Model Selection

The `fit_motif_auto()` function automatically fits all motif types and ranks them by goodness-of-fit criteria.

```{r auto_selection}
# Automated model selection
results <- fit_motif_auto(clay_profile, criterion = "AIC")
print(results)

# Plot the results
plot(results)
```

The function returns an object with:

- `fits`: All fitted motifs
- `stats`: Fit statistics (RMSE, AIC, BIC, R^2) for each model
- `ranking`: Models ranked by the chosen criterion
- `best`: Name of the best model

# Interpreting Results

Different criteria emphasize different aspects of model performance:

- **AIC/BIC**: Balance fit quality with model complexity (prefer lower values)
- **RMSE**: Absolute fit error (prefer lower values)  
- **R^2**: Proportion of variance explained (prefer higher values)

```{r interpretation}
# Compare fit statistics
print(results$stats)

# Best model parameters
best_params <- attr(results$fits[[results$best]], 'par')
cat("Best model:", results$best, "\n")
cat("Parameters:", paste(best_params, collapse = ", "), "\n")
```

# Working with Multiple Profiles

You can process multiple soil profiles simultaneously:

```{r multiple_profiles}
# Multiple profiles
multi_data <- data.frame(
  id = c("P1", "P1", "P1", "P2", "P2", "P2"),
  top = c(0, 20, 60, 0, 25, 70),
  bottom = c(20, 60, 100, 25, 70, 120),
  clay = c(12, 28, 38, 8, 22, 45)
)

# Prepare all profiles
all_profiles <- sm_prepare(multi_data, property = "clay", method = "step")

# Fit motifs for each profile
results_list <- lapply(all_profiles$clay, function(profile) {
  fit_motif_auto(profile, criterion = "AIC")
})

# Show results for each profile
for (profile_id in names(results_list)) {
  cat("\nProfile", profile_id, "- Best model:", results_list[[profile_id]]$best, "\n")
}
```

# Advanced Usage

## Custom Objective Functions

You can use different objective functions for optimization:

```{r custom_objective}
# Use sum of squared errors instead of RMSE
custom_fit <- sm_optim(clay_profile, c(50, 100), 
                      FUN = sm_shape_wetting_front,
                      OPTFUN = sm_optim_ssq)
```

## Handling Missing Data

The preprocessing functions handle missing values appropriately:

```{r missing_data}
# Data with missing values
missing_data <- data.frame(
  id = c("P1", "P1", "P1"),
  top = c(0, 30, 70),
  bottom = c(30, 70, 100),
  clay = c(15, NA, 40)  # Missing middle horizon
)

# Linear interpolation will handle the NA
missing_profile <- sm_prepare(missing_data, property = "clay", method = "linear")$clay$P1
plot(missing_profile, depths, ylim = c(200, 1), type = "l",
     main = "Profile with Missing Data (Linear Interpolation)")
```

This vignette covers the core workflow for soil motif analysis. For more details on specific motif types and their pedogenic interpretations, see the individual function documentation.